Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-30 _tex/paper2.tex
Built site for gh-pages

Before incorporating the first theoretical principles of Thurstone's
theory, it is essential to further define SCMs. SCMs are formal
mathematical models characterized by a set of \emph{endogenous}
variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
functions \(F\)
\citep{Pearl_2009, Pearl_et_al_2016, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly \citep{Pearl_2009}.
Lastly, the functions, referred to as \emph{structural equations},
express the endogenous variables as non-parametric functions of other
endogenous and exogenous variables. These functions use the symbol
`\(:=\)' to denote the asymmetrical causal dependence between variables
and the symbol `\(\:\bot\:\)' to represent \emph{d-separation}, a
concept akin to (conditional) independence.

SCM~\ref{fig-cj03_scm} presents the first theoretical principles
embedded in the conceptual-population model, which evaluates the impact
of different teaching methods on students' writing ability. This SCM
outlines the relationship between the conceptual-population outcome
\((O^{cp}_{iahbjk})\) and several related variables. The subscripts
\(i\) and \(h\) identify the students who authored the texts (i.e., the
individuals). The indices \(a\) and \(b\) represent the texts under
comparison (i.e., the stimuli). The index \(j\) indicates the judge
conducting the comparison, while the index \(k\) captures \emph{repeated
measures designs} \citep[pp.~366-376]{Lawson_2015}, accounting for
experimental conditions where a judge compares the same pair of stimuli
multiple times. Thus, the indexing system supports comparisons between
different texts written by the same student \((i = h;\) \(a \neq b)\)
and between texts written by distinct students \((i \neq h;\) where
\(a = b\) is permitted\()\), each judged once or repeatedly by all
judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K;\) where \(n_{J}>1\)
and \(n_{K}\geq1)\). However, it excludes cases where a judge compares a
student's text to itself, whether once or multiple times \((i = h;\)
\(a = b;\) \(j = 1,\dots,n_{J};\) \(k = 1,\dots,n_{K};\) where
\(n_{J}>1\) and \(n_{K}\geq1)\), as such comparison lacks practical
relevance within the CJ framework. Here, \(n_{J}\) indicates the total
number of judges, and \(n_{K}\) denotes the number of repeated judgments
each judge performs.

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ia}, T_{hb})\) and their discriminal
difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, the SCM incorporates a key design feature of CJ
experiments: the judges' biases \((B_{kj})\). This extension builds on
the arguments presented in Section~\ref{sec-theory-issue1b}, contending
that the discriminal difference becomes an observable outcome only
through judges' perceptions. Given that such perceptions may be
imperfect--and that each judge may carry some degree of bias
\citep[see][]{Pollitt_et_al_2003, vanDaal_et_al_2016}--it is reasonable
that judges' perceptions (bias) should be treated as an integral
component of the CJ system from the outset, as this leads to a more
accurate representation of the data-generating process underlying the
pairwise comparisons. All these variables, together with the outcome
\((O_{iahbjk})\), constitute the preliminary set of endogenous
variables, \(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\).
Finally, the SCM presents the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

It is easier to visualize the structure of these vectors and matrices by
considering an example where \(n_{I} = 5\), \(n_{A} = 2\),
\(n_{J} = 3\), and \(n_{K} = 3\). In this simple case, the vectors and
matrices described in equations (\ref{eq-mat11}) and (\ref{eq-mat12})
take the following form:
\begin{equation}\phantomsection\label{eq-mat13}{
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 2 & 1 & 1 \\
1 & 1 & 5 & 2 & 1 & 2 \\
1 & 1 & 5 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 
\end{bmatrix}
}\end{equation}

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citep{Fox_2010, vanderLinden_et_al_2017_I}, the conceptual-population
model integrates two \emph{hierarchical structural components} to
examine how different teaching methods influence students' writing
ability. Each structural component defines how observed or latent
variables affect the primary latent variable of interest
\citep{Everitt_et_al_2010}. The model's hierarchical design allows
researchers to formulate and test hypotheses that account for both the
nested structure of stimuli and the uncertainties inherent in trait
estimation (see Section~\ref{sec-theory-issue1b} and
Section~\ref{sec-theory-issue2} for further discussion of these
considerations).

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{\emph{Relevant variables} are
  those that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. Refer also to footnote 1.} student-related
variables \((X_{I})\), such as teaching method, and students'
idiosyncratic errors \(e_{I}\) causally influence the latent variable
representing students' writing-quality trait \((T_{I})\). The error term
\(e_{I}\) captures variations in students' traits unexplained by
\(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and
\(q_{I}\) independent columns (variables), and both \(e_{I}\) and
\(T_{I}\) are latent vectors of length \(n_{I}\). Additionally, this
branch shows how \(T_{I}\), along with \emph{relevant} \footnote{refer
  to footnote 2.} text-related variables \((X_{IA})\) (e.g., text
length), and texts' idiosyncratic errors \(e_{IA}\) causally influence
the texts' written-quality trait \((T_{IA})\), the first primary latent
variable of interest. The error term \(e_{IA}\) captures variations in
the texts' traits that remain unexplained by \(T_{I}\) or \(X_{IA}\).
Here, \(X_{IA}\) is an observed matrix with dimensions
\(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) independent columns
(variables), while \(e_{IA}\) and \(T_{IA}\) are latent matrices with
\(n_{I}\) rows and \(n_{A}\) columns.

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 2.}
judge-related variables \((Z_{J})\), such as judgment expertise, and
judges' idiosyncratic errors \(e_{J}\) causally influence the latent
variable representing judges' bias \((B_{J})\). The error \(e_{J}\)
captures variations in judges' bias unexplained by \(Z_{J}\). Here,
\(Z_{J}\) is an observed matrix with \(n_{J}\) rows and \(q_{J}\)
independent columns (variables), and both \(e_{J}\) and \(B_{J}\) are
latent vectors of length \(n_{J}\). Furthermore, the branch shows how
\(B_{J}\), along with \emph{relevant} \footnote{refer to footnote 2.}
judgment-related variables \((Z_{JK})\) (e.g., the number of judgments a
judge makes), and judgments' idiosyncratic errors \(e_{JK}\) causally
influence the judges' biases associated with each text \(B_{JK}\), the
second primary latent variable of interest. The error \(e_{JK}\)
captures variations in judgments unexplained by \(B_{J}\) or \(Z_{JK}\).
Here, \(Z_{JK}\) is an observed matrix with dimension
\(n_{J} \cdot n_{K}\) rows and \(q_{JK}\) independent columns
(variables), while \(e_{JK}\) and \(B_{JK}\) are latent latent matrices
with \(n_{J}\) rows and \(n_{K}\) columns

Notably, all variables and functions shown in SCM~\ref{fig-cj09_scm} and
DAG~\ref{fig-cj09_dag} are part of the set of endogenous variables
\(V\), structural equations \(F\), and exogenous variables \(E\) for the
conceptual-population model. Additionally, the figures demonstrate that
all exogenous variables are independent of one another, as indicated by
the relationships \(e_{IA} \:\bot\:\{ e_{I}, e_{JK}, e_{J} \}\),
\(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\).

@@

Equation (\ref{eq-mat24}) shows that missing observations in the vectors
\(S_{I}\) and \(S_{J}\)--which represent unsampled students and
judges--directly determine which observations are missing in \(S_{IA}\)
and \(S_{JK}\). In other words, researchers can only observe texts and
judgments from students and judges initially included in the sample. The
equation also shows that the sum of observed elements in \(S_{I}\)
equals the number of sampled students \((n^{s}_{I})\) and that a similar
sum in vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\).
Conversely, the sum of observed elements in \(S_{IA}\) represents the
total sampled texts across all sampled students
\((n^{s}_{I} \cdot n^{s}_{A})\), while a similar sum in vector
\(S_{JK}\) represents the total sampled repeated judgments across all
sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\). Notice that because the
design systematically excludes every third repeated judgment,
researchers can also express \(S_{JK}\) using \(n_{K} = n^{s}_{K} = 2\).

@@

With the definition of \(S\), we incorporate the sample mechanism into
the conceptual-population model. Following the convention of
\citet{McElreath_2020} and \citet{Deffner_et_al_2022},
DAG~\ref{fig-cj14_dag} represents the conceptual-population outcome
\(O^{cp}_{R}\) as unobserved, emphasizing that researchers cannot
directly access this outcome due to the sampling mechanism. The DAG also
depicts the \emph{sample design} vector \(S\) as a causal factor
influencing the sample-comparison outcome \(O^{sc}_{R}\). A square
encloses this vector, indicating that it is a conditioned variable.
Here, \emph{conditioning} means that researchers restrict their focus to
the elements of \(O^{cp}_{R}\) that satisfy \(s_{(i,a,h,b,j,k)}=1\)
\citep{Neal_2020, McElreath_2020}. Essentially, \(S\) serves as the
vector selecting \emph{all repeated judgments a subset of judges makes
for the subset of stimuli produced by the sampled individuals}. This
process generates the sample-comparison outcome \(O^{sc}_{R}\).

@@

The DAG~\ref{fig-cj14_dag} also integrates the \emph{comparison
mechanism} \(C\) into the conceptual-population model, ultimately
deriving the outcome \(O_{R}\). It shows the sample-comparison outcome
as unobserved, highlighting that researchers cannot directly access to
\(O^{sc}_{R}\) due to the comparison mechanism. Moreover, the DAG
depicts \(C\) as a causal factor influencing \(O_{R}\), enclosing it in
a square to indicate that it is a conditioned variable. Specifically,
\(C\) determines \emph{which repeated judgments the subset of judges
makes for the subset of stimuli produced by the sampled individuals}. In
essence, \(C\) reflects the assumption that judges \emph{do not} perform
all possible repeated judgments for the selected stimuli. Instead, they
carry out a ``sufficient number of comparisons'' \(n_{C}\) to allow for
accurate estimation of the proportion \(P(B>A)\) for each stimulus pair
\citep[pp.~267]{Thurstone_1927b}.

Notably, DAG~\ref{fig-cj14_dag} also shows that \(C\) is independent of
all other variables in the model. This independence implies that the
conceptual model represented by the DAG applies exclusively to Random
Allocation Comparative Designs \citep{Bramley_2015} or Incomplete Block
Designs \citep{Lawson_2015}. In these designs, every repeated judgment
has an equal probability of being included in the sample.

@@

\section{From SCM to a statistical model}\label{sec-statistical}

@@

Each CPD in equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline in the SCM~\ref{fig-cj16_scm}, the probabilistic model
\ref{fig-cj16_scm}, and the statistical model \ref{fig-cj16_scm}. First,
we assume that \(O_{R}\) follows a Bernoulli distribution\footnote{The
  binomial distribution--including its special case, the Bernoulli
  distribution-- represent a maximum entropy distribution for binary
  events \citep[pp.~34]{McElreath_2020}. This means that when only two
  un-ordered outcomes are possible, and the expected numbers of each
  type of event are assumed to be constant, then the binomial
  distribution is the one that is most consistent with these constraints
  \citep[pp.~310]{McElreath_2020}. For a detailed discussion of the
  binomial as a maximum entropy distribution, see
  \citet[chap.~10.1.2]{McElreath_2020}.}, reflecting the binary nature
of CJ outcomes. Furthermore, following the conventions of Generalized
Linear Latent and Mixed Models (GLLAMMs)
\citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\).

@@

Note that while the joint distribution in Equation (\ref{eq-mat51})
includes the probability distributions of the sampling and comparison
mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
\(P(Z_{J})\)--we omit these probabilities from the models presented in
Figure~\ref{fig-cj16}. We make this omission because, although these
distributions contribute to the overall joint distribution of the data,
they do not influence the relationship between \(O_{R}\) and \(D_{R}\).
This independence stems from the random selection designs that govern
\(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
\(Z_{J}\)\footnote{Randomization ensures that data---and, by extension,
  an estimator---satisfies several key identification properties, such
  as common support, no interference, and consistency. The most critical
  property, however, is the elimination of confounding.
  \emph{Confounding} occurs when an external variable, such as
  \(X_{I}\), simultaneously influences both the outcome (e.g.,
  \(O_{R}\)) and a variable of interest (e.g., \(X_{I}\) or \(S\)),
  resulting in spurious associations between the two
  \citep{Everitt_et_al_2010}. Randomization addresses this issue by
  effectively decoupling the association between the variable of
  interest and any other variable, except for the outcome itself. For a
  more detailed discussion on the benefits of randomization, see
  \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, \citet{Neal_2020}, and
  \citet{Hernan_et_al_2020}.}.

Second, we define \(D_{R}\) as the difference between the discriminal
processes of two texts, \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), plus the
corresponding repeated judge biases, \(B_{JK}[j, k]\). The discriminal
processes represent the underlying written-quality trait of the texts.
Note that if we assume that \(B_{JK}[j,k]\) reflects the difference in
stimulus-specific biases, i.e.,
\(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can re-write the
discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat52}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation} This formulation reveals that the discriminal difference
captures a \emph{pure interaction effect}, in which neither the texts'
discriminal processes nor the judges' biases alone determine the
outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
terms, the discriminal processes of the stimuli become an observable
outcome only through the judges' perceptions (biases). It is important
to clarify that the square brackets in \(D_{R}\) merely indicate the
relevant indices for each trait vector, without implying any subsetting
of the data.

>>>>

Before incorporating the first theoretical principles of Thurstone's
theory, it is essential to further define SCMs. SCMs are formal
mathematical models characterized by a set of \emph{endogenous}
variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
functions \(F\)
\citep{Pearl_2009, Pearl_et_al_2016, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly \citep{Pearl_2009}.
Lastly, the functions, referred to as \emph{structural equations},
express the endogenous variables as non-parametric functions of other
endogenous and exogenous variables. These functions use the symbol
`\(:=\)' to denote the asymmetrical causal dependence between variables
and the symbol `\(\:\bot\:\)' to represent \emph{d-separation}, a
concept akin to statistical (conditional) independence.

SCM~\ref{fig-cj03_scm} presents the first theoretical principles
embedded in the conceptual-population model, which evaluates the impact
of different teaching methods on students' writing ability. This SCM
outlines the relationship between the conceptual-population outcome
\((O^{cp}_{iahbjk})\) and several related variables. The subscripts
\(i\) and \(h\) identify the students who authored the texts (i.e., the
individuals). The indices \(a\) and \(b\) represent the texts under
comparison (i.e., the stimuli). The index \(j\) indicates the judge
conducting the comparison, while the index \(k\) accounts for
experimental conditions where a judge compares the same pair of stimuli
multiple times, i.e., a \emph{repeated measures designs}
\citep[pp.~366-376]{Lawson_2015}. Thus, the indexing system supports
comparisons between different texts written by the same student
\((i = h;\) \(a \neq b)\) and between texts written by distinct students
\((i \neq h;\) where \(a = b\) is permitted\()\), each compared once or
repeatedly by all judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K;\)
where \(n_{J}>1\) and \(n_{K}\geq1)\). However, it excludes cases where
a judge compares a student's text to itself, whether once or multiple
times \((i = h;\) \(a = b;\) \(j = 1,\dots,n_{J};\)
\(k = 1,\dots,n_{K};\) where \(n_{J}>1\) and \(n_{K}\geq1)\), as such
comparison lacks practical relevance within the CJ framework. Here,
\(n_{J}\) indicates the total number of judges, and \(n_{K}\) denotes
the number of repeated judgments each judge performs.

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ia}, T_{hb})\) and their discriminal
difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, the SCM incorporates a key design feature of CJ
experiments: the judges' biases \((B_{kj})\). This extension builds on
the arguments presented in Section~\ref{sec-theory-issue1b}, contending
that the discriminal difference becomes an observable outcome only
through judges' perceptions. Given that such perceptions may be
imperfect--and that each judge may carry some degree of bias
\citep[see][]{Pollitt_et_al_2003, vanDaal_et_al_2016}--it is reasonable
that judges' perceptions (bias) should be treated as an integral
component of the CJ system from the outset, as this leads to a more
accurate representation of the data-generating process underlying the
pairwise comparisons. All these variables constitute the preliminary set
of endogenous variables,
\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\), while the
preliminary set of structural equations, \(F = { f_{O}, f_{D} }\),
captures the non-parametric dependencies among them.

@@

It is easier to visualize the structure of these vectors and matrices by
considering an example. Assuming \(n_{I} = 5\), \(n_{A} = 2\),
\(n_{J} = 3\), and \(n_{K} = 3\), the vectors and matrices described in
equations (\ref{eq-mat11}) and (\ref{eq-mat12}) take the following form:
\begin{equation}\phantomsection\label{eq-mat13}{
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 2 & 1 & 1 \\
1 & 1 & 5 & 2 & 1 & 2 \\
1 & 1 & 5 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 
\end{bmatrix}
}\end{equation}

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citep{Fox_2010, vanderLinden_et_al_2017_I}, the conceptual-population
model integrates two \emph{hierarchical structural components} to
examine how different teaching methods influence students' writing
ability. Each structural component defines how observed or latent
variables affect the primary latent variable of interest
\citep{Everitt_et_al_2010}. The model's hierarchical design allows
researchers to formulate and test hypotheses that account for both the
nested structure of stimuli and the uncertainties inherent in trait
estimation (see Section~\ref{sec-theory-issue1b} and
Section~\ref{sec-theory-issue2} for a discussion of these
considerations).

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{\emph{Relevant variables} are
  those that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. Refer also to footnote 1.} student-related
variables \(X_{I}\), such as teaching method, and students'
idiosyncratic errors \(e_{I}\) causally influence the latent variable
representing students' writing-quality trait \(T_{I}\). The error term
\(e_{I}\) captures variations in students' traits unexplained by
\(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and
\(q_{I}\) independent columns (variables), and both \(e_{I}\) and
\(T_{I}\) are latent vectors of length \(n_{I}\). Additionally, this
branch shows how \(T_{I}\), along with \emph{relevant} \footnote{refer
  to footnote 2.} text-related variables \(X_{IA}\) (e.g., text length),
and texts' idiosyncratic errors \(e_{IA}\) causally influence the texts'
written-quality trait \(T_{IA}\), the first primary latent variable of
interest. The error term \(e_{IA}\) captures variations in the texts'
traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). Here,
\(X_{IA}\) is an observed matrix with dimensions \(n_{I} \cdot n_{A}\)
rows and \(q_{IA}\) independent columns (variables), while \(e_{IA}\)
and \(T_{IA}\) are latent matrices with \(n_{I}\) rows and \(n_{A}\)
columns.

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 2.}
judge-related variables \(Z_{J}\), such as judgment expertise, and
judges' idiosyncratic errors \(e_{J}\) causally influence the latent
variable representing judges' bias \(B_{J}\). The error \(e_{J}\)
captures variations in judges' bias unexplained by \(Z_{J}\). Here,
\(Z_{J}\) is an observed matrix with \(n_{J}\) rows and \(q_{J}\)
independent columns (variables), and both \(e_{J}\) and \(B_{J}\) are
latent vectors of length \(n_{J}\). Furthermore, the branch shows how
\(B_{J}\), along with \emph{relevant} \footnote{refer to footnote 2.}
judgment-related variables \(Z_{JK}\) (e.g., the number of judgments a
judge makes), and judgments' idiosyncratic errors \(e_{JK}\) causally
influence the judges' biases associated with each text \(B_{JK}\), the
second primary latent variable of interest. The error \(e_{JK}\)
captures variations in judgments unexplained by \(B_{J}\) or \(Z_{JK}\).
Here, \(Z_{JK}\) is an observed matrix with dimension
\(n_{J} \cdot n_{K}\) rows and \(q_{JK}\) independent columns
(variables), while \(e_{JK}\) and \(B_{JK}\) are latent latent matrices
with \(n_{J}\) rows and \(n_{K}\) columns

Notably, all variables and functions shown in SCM~\ref{fig-cj09_scm} and
DAG~\ref{fig-cj09_dag} are part of the set of endogenous variables
\(V\), structural equations \(F\), and exogenous variables \(E\) for the
conceptual-population model. Additionally, the figures demonstrate that
all exogenous variables are independent of one another, as indicated by
the relationships \(e_{IA} \:\bot\:\{ e_{I}, e_{JK}, e_{J} \}\),
\(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\) and
the absence of connecting arrows.

@@

Equation (\ref{eq-mat24}) shows that missing observations in the vectors
\(S_{I}\) and \(S_{J}\)--which represent unsampled students and
judges--directly determine which observations are missing in \(S_{IA}\)
and \(S_{JK}\). In other words, researchers can only observe texts and
judgments from students and judges initially included in the sample. The
equation also shows that the sum of observed elements in \(S_{I}\)
equals the number of sampled students \((n^{s}_{I})\) and that a similar
sum in vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\).
Conversely, the sum of observed elements in \(S_{IA}\) represents the
total sampled texts across all sampled students
\((n^{s}_{I} \cdot n^{s}_{A})\), while a similar sum in vector
\(S_{JK}\) represents the total sampled repeated judgments across all
sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\). Notice that in this
example, because the design systematically excludes every third repeated
judgment, researchers can also express \(S_{JK}\) using
\(n_{K} = n^{s}_{K} = 2\).

@@

With the definition of \(S\), we incorporate the sample mechanism into
the conceptual-population model. Following the convention of
\citet{McElreath_2020} and \citet{Deffner_et_al_2022},
DAG~\ref{fig-cj14_dag} represents the conceptual-population outcome
\(O^{cp}_{R}\) as unobserved, emphasizing that researchers cannot
directly access this outcome due to the sampling mechanism. The DAG also
depicts the \emph{sample design} vector \(S\) as a causal factor
influencing the sample-comparison outcome \(O^{sc}_{R}\). A square
encloses this vector, indicating that it is a conditioned variable.
Here, \emph{conditioning} means that researchers restrict their focus to
the elements of \(O^{cp}_{R}\) that satisfy \(s_{(i,a,h,b,j,k)}=1\)
\citep{Neal_2020, McElreath_2020}. Essentially, \(S\) serves as the
vector selecting \emph{all repeated judgments a subset of judges makes
for the subset of stimuli produced by the sampled individuals}.

@@

The DAG~\ref{fig-cj14_dag} also integrates the \emph{comparison
mechanism} \(C\) into the conceptual-population model. It shows the
sample-comparison outcome as unobserved, highlighting that researchers
cannot directly access to \(O^{sc}_{R}\) due to the comparison
mechanism. Moreover, the DAG depicts \(C\) as a causal factor
influencing \(O_{R}\), enclosing it in a square to indicate that it is a
conditioned variable. Specifically, \(C\) determines \emph{which
repeated judgments the subset of judges makes for the subset of stimuli
produced by the sampled individuals}. In essence, \(C\) reflects the
assumption that judges \emph{do not} perform all possible repeated
judgments for the selected stimuli. Instead, they carry out a
``sufficient number of comparisons'' \(n_{C}\) to allow the accurate
estimation of the proportion \(P(B>A)\) for each stimulus pair
\citep[pp.~267]{Thurstone_1927b}.

Notably, DAG~\ref{fig-cj14_dag} also shows that \(C\) is independent of
all other variables in the model. This independence implies that the
conceptual model represented by the DAG applies exclusively to Random
Allocation Comparative Designs \citep{Bramley_2015}, or Incomplete Block
Designs \citep{Lawson_2015}, where every repeated judgment has an equal
probability of being included in the sample.

@@

\section{From SCM to statistical model}\label{sec-statistical}

@@

Each CPD in equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline in the SCM~\ref{fig-cj16_scm}, the probabilistic model
\ref{fig-cj16_prob}, and the statistical model \ref{fig-cj16_stat}.
First, we assume that \(O_{R}\) follows a Bernoulli
distribution\footnote{The binomial distribution--including its special
  case, the Bernoulli distribution--represent a maximum entropy
  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
  means that when only two un-ordered outcomes are possible, and the
  expected numbers of each type of event are assumed to be constant,
  then the binomial distribution is the one that is most consistent with
  these constraints \citep[pp.~310]{McElreath_2020}. For a detailed
  discussion of the binomial as a maximum entropy distribution, see
  \citet[sec.~10.1.2]{McElreath_2020}.}, reflecting the binary nature of
CJ outcomes. Furthermore, following the conventions of Generalized
Linear Latent and Mixed Models (GLLAMMs)
\citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\).

@@

Note that while the joint distribution in Equation (\ref{eq-mat51})
includes the probability distributions of the sampling and comparison
mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
\(P(Z_{J})\)--we omit these probabilities from the models presented in
Figure~\ref{fig-cj16}. We make this omission because, although these
distributions contribute to the overall joint distribution of the data,
they do not influence the relationship between \(O_{R}\) and \(D_{R}\).
This independence stems from the random selection designs that govern
\(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
\(Z_{J}\)\footnote{Randomization ensures that data---and, by extension,
  an estimator---satisfies several key identification properties, such
  as common support, no interference, and consistency. The most critical
  property, however, is the elimination of confounding.
  \emph{Confounding} occurs when an external variable, such as
  \(X_{I}\), simultaneously influences both the outcome (e.g.,
  \(O_{R}\)) and a variable of interest (e.g., \(X_{I}\) or \(S\)),
  resulting in spurious associations between the two
  \citep{Everitt_et_al_2010}. Randomization ensure the absence of
  confounding by effectively decoupling the association between the
  variable of interest and any other variable, except for the outcome
  itself. For a more detailed discussion on the benefits of
  randomization, see \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
  \citet{Neal_2020}, and \citet{Hernan_et_al_2020}.}.

Second, we define \(D_{R}\) as the difference between the discriminal
processes of two texts--\(T_{IA}[i, a]\) and
\(T_{IA}[h, b]\)--representing the underlying written-quality trait of
the texts, plus the corresponding repeated judge biases,
\(B_{JK}[j, k]\). Note that if we assume that \(B_{JK}[j,k]\) reflects
the difference in stimulus-specific biases, i.e.,
\(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can re-write the
discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat52}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation} This formulation reveals that the discriminal difference
captures a \emph{pure interaction effect}, in which neither the texts'
discriminal processes nor the judges' biases alone determine the
outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
terms, the discriminal processes of the stimuli become an observable
outcome only through the judges' perceptions (biases). To clarify, the
square brackets in \(D_{R}\) simply indicate the relevant indices for
each trait vector and do not imply any subsetting of the data.
